How environmental drivers of spatial synchrony interact

### 2023-02-07 Paper/Paper.Rnw
almost final version, sent to coauthors for final look

%Intro to Moran effects and history of the early years where people had trouble identifying them
Spatial autocorrelations in fluctuating environmental drivers of populations are now known to be one of the most important
causes of synchrony, through a process called the Moran effect. But this effect was originally proposed 
theoretically \citep{Moran1953}, and decades passed during which it was 
considered difficult to distinguish this cause of synchrony from others, except under special circumstances.
Synchrony has long been thought to have three causes: dispersal, and trophic interactions with a synchronous or
mobile species, in addition to the Moran effect \citep{Liebhold2004}. But using common past approaches
examining declines in population correlations with distance, statistical patterns of synchrony produced
by each of these mechanisms are quite similar \citep{Ranta1999,Abbott2007,Liebhold2004,Walter2017}. So examination
of such patterns provides little or no traction for performing inferences about causes of synchrony. 
Early papers demonstrating the importance of Moran effects resorted to special cases where dispersal was 
impossible and predators were absent \citep{Grenfell1998,Tedesco2004}. Though controlled experiments revealed that all
three mechanisms could be involved in synchrony [e.g., \citealp{Vasseur2009}], the problem of determining
the relative contribution of each mechanism to synchrony observed in field systems was considered a challenge 
for decades.

%Recent statistical explorations of Moran effects

%Intro to interactions
Two recent papers have documented that Moran effects of distinct 
environmental drivers can, in fact, interact, either synergistically or antagonistically,
so that total population synchrony can be either greater than or less than what synchrony would be if the distinct Moran 
drivers operated independently \citep{Sheppard2019,CastoraniSheppardKelp}. 
\cite{Sheppard2019} showed that long-timescale ($>4$y period) synchrony in a chlorophyll density index 
in the seas around the United Kingdom is substantially augmented by interactions between environmental 
drivers of synchrony and drivers linked with consumption by a copepod consumer.
\cite{CastoraniSheppardKelp} showed that nutrient dynamics and wave disturbance, two Moran drivers
of synchrony in giant kelp populations on the California (CA) coast, interact either synergistically 
or destructively, depending on which portion of the CA coast is examined and on 
the timescale of analysis. 
The main purpose of this study is to establish a general theory of the mechanism
of interacting Moran effects and to use examples and applications of the theory to provide 
ecological intuition for how the mechanism of interacting Moran effects works. Our goals are
distinct from those of \cite{Kendall2000}, who examined interactions between Moran and dispersal
causes of synchrony.

@@

Our formal theory requires a conceptual understanding of the \emph{spectrum} of an environmental 
or population time series, as well as of the \emph{cospectrum} and \emph{cross spectrum} of two time series,
so we briefly introduce these concepts.
If $w_i(t)$ is a stochastic process or time series measured at location $i$ (e.g., the population density time series
of a species of interest in that location), the spectrum $S_{w_i w_i}(f)$ is a function
of frequency, $f$. For a periodic ocillation, $f$ can be defined as one over 
the timescale, or period, $\sigma$, of the oscillation. 
The spectrum $S_{w_i w_i}(f)$ is larger for frequencies at which $w_i(t)$ oscillates more. So, for example, a 
population $w_i(t)$ that shows strong oscillatory dynamics of $5$-year period will have a large value of 
$S_{w_i w_i}(f)$ for $f=\frac{1}{5} \text{ y}^{-1}$. The spectrum separates the overall variance of a time series into
contributions which occur at different frequencies, in the sense that an integral of $S_{w_i w_i}(f)$ across all 
frequencies equals $\var(w_i)$. In a similar way, the cospectrum of two time series, $w_i(t)$ and $w_j(t)$,
is a function of $f$ that takes large values at frequencies for which oscillations in $w_i(t)$ and 
$w_j(t)$ are both strong and strongly correlated, i.e., they have the same or similar phase. 
Here, $j$ refers to another location where measurements were taken. The cospectrum is 
the real part of the cross spectrum, $S_{w_i w_j}(f)$, which is a complex-valued function of frequency.
The cross-spectrum takes large-magnitude values at frequencies, $f$, for which the oscillatory components
of $w_i(t)$ and $w_j(t)$ are strong and in a consistent phase relationship to each other; 
and the complex phase of the cross spectrum at $f$ then quantifies
that relationship. The notation $S_{ww}(f)$ refers to the \emph{spectral matrix}, 
which has $ij^{\text{th}}$ entry $S_{w_i w_j}(f)$. Spectral methods are standard  
[e.g., \cite{Vasseur2007,Reuman2008,Defriez2017}], and many background references are available 
[e.g., \cite{Brillinger2001}].

@@

Our population model is 
\begin{linenomath}
\begin{align}
w_i(t) &= b_1 w_i(t-1)+ \cdots + b_n w_i(t-n) \label{main_model_1} \\ 
&+ p_0^{(1)} \epsilon_i^{(1)}(t)+ \cdots + p_{m_1}^{(1)} \epsilon_i^{(1)}(t-m_1) \label{main_model_2} \\
&+ p_0^{(2)} \epsilon_i^{(2)}(t)+ \cdots + p_{m_2}^{(2)} \epsilon_i^{(2)}(t-m_2) \label{main_model_3} \\
&+ \delta_i(t), \label{main_model_4}
\end{align}
\end{linenomath}
where $w_i(t)$ is an index of the population in location $i=1,\ldots,N$ at time
$t$, and $\epsilon^{(1)}=(\epsilon_1^{(1)},\ldots,\epsilon_N^{(1)})$,
$\epsilon^{(2)}=(\epsilon_1^{(2)},\ldots,\epsilon_N^{(2)})$ and $\delta=(\delta_1,\ldots,\delta_N)$
are environmental processes at the same locations. The processes $\epsilon^{(1)}$ and $\epsilon^{(2)}$ are
taken to be measured, whereas $\delta$ represents the 
aggregate influence of unmeasured processes. We assume that the combined process 
$(\epsilon_1^{(1)},\ldots,\epsilon_N^{(1)},\epsilon_1^{(2)},\ldots,\epsilon_N^{(2)},\delta_1,\ldots,\delta_N)$ is an ergodic second-order stationary stochastic process \citep{Brillinger2001}
with expected values of its components equal to $0$. We make additional mild
regularity assumptions for model stability 
(SI section \ref{SM-sect:model_spec}). This model can be seen as 
a linearization of a very general dynamical model, influenced by ``weak noise'' [see, e.g.,
SI section S1.2 of \cite{Desharnais2018} and SI section S1 of \cite{Walter2017}].
Linearization and ``weak noise'' assumptions have been commonly adopted to make
theoretical progress in ecology, and it has been demonstrated that results based on a
weak noise assumption often hold for noise which is fairly strong \citep{Nisbet1977,Reuman2006,Desharnais2018}.
See \cite{Brillinger2001} for background on stochastic processes.

@@

%*DONE relationship to earlier papers comparing synchrony of drivers and pops
It has been a frequent topic of research why populations are often less synchronous, or sychronous over a smaller
spatial extent, than might be expected given the strength and extent of 
synchrony of an environmental driver \citep{Herfindal2022}. Our new mechanism
of interacting Moran effects provides both a new means by which populations may be less synchronous than 
population drivers; and also a new means by which populations can be more 
synchronous than environmental drivers. 
Previously known mechanisms by which populations can be less synchronous than environmental drivers 
include demographic stochasticity and measurement error. 
Antagonistic interactions between Moran drivers may be a common and previously
unrecognized additional mechanism contributing to this discrepancy. On the other hand,
two recent papers descibed ``enhanced Moran effects'' by which specific 
patterns of temporal autocorrelation
in Moran drivers can theoretically cause greater synchrony in populations 
than in drivers \citep{Massie2015,Desharnais2018}.
Synergistic interaction between Moran drivers are another mechanism by which populations can
be more synchronous than expected. Although we considered only two drivers in 
our theory and examples, populations may, in principle, depend on many environmental
factors, all related to a large-scale climatic phenomenon such as 
ENSO and hence related to each other. Future work should consider whether and when such scenarios can lead to elevated synchrony.

@@

% %*DONE Revisit the kendall paper, briefly. Talk about possible three-way interactions.
% %***DAN: Could def put this in sup mat if we're cutting, or drop it because it's a bit lame
% \cite{Kendall2000} considered interactions between mechanisms of synchrony. Their research 
% questions were distinct from ours, but combining their viewpoints and ours may lead to valuable
% future work. \cite{Kendall2000} considered interactions between dispersal and Moran causes of 
% synchrony, whereas we analyzed interactions between two Moran drivers, a distinct context.
% But it seems likely that numuerous ecological systems involve important dispersal effects on 
% synchrony as well as influences of at least two Moran effects. It may be necessary in future
% work to consider three-way interactions between dispersal and two Moran drivers and their 
% influence on synchrony. Dispersal can readily be added to our modelling framework.
% \cite{Desharnais2018} performed a spectral analysis on a model similar to ours
% which included dispersal but which only considered two habitat patches and one environmental driver.

>>>>

%Intro to Moran effects and history of the early years where people had trouble identifying them
Correlations between weather time series measured in different locations can 
induce synchrony between populations in those locations if the weather variables influence
population processes. This mechanism, called the Moran effect, is now known to be one of the most
important causes of synchrony. But the mechanism was originally proposed 
theoretically \citep{Moran1953}, and decades passed during which it was 
considered difficult to distinguish this potential cause of synchrony from others in field systems.
Synchrony has long been thought to have three causes: dispersal, and trophic interactions with a synchronous or
mobile species, in addition to the Moran effect \citep{Liebhold2004}. But using common past statistical approaches
which focussed on declines in population correlations with distance, patterns of synchrony produced
by each of these mechanisms are quite similar \citep{Ranta1999,Abbott2007,Liebhold2004,Walter2017}; so examination
of such patterns provides little or no traction for inferring the causes of synchrony. 
Early papers demonstrating the Moran effect mechanism resorted to special cases where dispersal was 
impossible and predators were absent \citep{Grenfell1998,Tedesco2004}. Controlled experiments also confirmed that all
three theorized mechanisms could be involved in synchrony [e.g., \citealp{Vasseur2009}]. But the problem of inferring
the relative contribution of the three mechanisms in field systems was considered a challenge 
until recently.

%Recent statistical explorations of Moran effects
Recent research has provided new statistical viewpoints which have, when sufficient data are available, 
essentially solved the problem of inference
of the causes of synchrony, and the research has revealed the broad importance of Moran effects. 
%geographic approaches
Approaches based on formal statistical comparisons between population synchrony maps and maps of 
environmental correlations produced inferences that precipitation and temperature Moran effects are important
causes of synchrony of terrestrial \citep{Defriez2017b} and marine \citep{Defriez2017} primary productivity. 
Another geographic approach, based on multiple regression of distance matrices (MRM), was used in an early paper
to infer that a precipitation Moran effect is a cause of synchrony for the spongy moth \citep{Haynes2013},
an invasive forest pest whose name was recently changed from ``gypsy moth.'' 
Geographic approaches to identifying causes of synchrony, often MRM approaches 
specifically \citep{Koenig2017,Walter2020,Bogdziewicz2021}, 
have become widespread, and the geography of synchrony has become a mainstream area of study \citep{Walter2017}.
The approach of \cite{Gouveia2016}, called ``noncentered local indicators of spatial association'' (ncLISA), is another 
geographic approach; as is the ``fluvial synchrogram'' of \cite{Larsen2021}, a tool for studying synchrony in 
river networks.
MRM approaches have also been used to identify dispersal as a cause of synchrony \citep{Anderson2018},
sometimes combined with genetic methods \cite{Haynes2022}.
%Time and timescale methods
Another class of methods exploit the time and timescale structure of synchrony.
Using a method that examined the match between temporal variations in population and environmental synchrony, 
\cite{Allstadt2015} confirmed the importance of a precipitation Moran effect for synchrony of spongy moth.
A suite of wavelet and related Fourier techniques have been developed that can comprehensively describe the time and timescale
structure of synchrony, identify Moran drivers of synchrony opperating in distinct timescale bands, and apportion fractions
of observed synchrony to respective Moran drivers \citep{Sheppard2016,Sheppard2017,Desharnais2018,Sheppard2019,wsyn,Anderson2021}.
The methods have been used, for instance, to provide one of the first clear demonstrations that changes in 
Moran effects are another of the impacts of climatic changes \citep{Sheppard2016,Sheppard2019}; and to illustrate the importance of the
timescale structure of environmental covariation for altering Moran effects \citep{Desharnais2018}.
These methods have since been used to identify Moran effects in several additional systems
(e.g., \citealp{Anderson2019,Walter2019,Walter2019,GarciaCarreras2022,CastoraniSheppardKelp}).

%Intro to interactions
The techniques reviewed above have made it possible for several recent papers to identify cases in which two or more
distinct Moran drivers opperate simultaneously on the same populations 
(e.g., \citealp{Defriez2017b,Haynes2019,Walter2019,Walter2020,Anderson2021}).
In fact, two recent papers have documented that Moran effects of distinct 
environmental drivers can interact, either synergistically or antagonistically,
so that total population synchrony can be either greater than or less than what synchrony would be if the distinct Moran 
drivers operated independently \citep{Sheppard2019,CastoraniSheppardKelp}. 
\cite{Sheppard2019} showed that long-timescale ($>4$y period) synchrony in a chlorophyll density index 
in the seas around the United Kingdom is substantially augmented by interactions between environmental 
drivers of synchrony and drivers linked with consumption by a copepod consumer.
\cite{CastoraniSheppardKelp} showed that nutrient dynamics and wave disturbance, two Moran drivers
of synchrony in giant kelp populations on the California (CA) coast, interact either synergistically 
or destructively, depending on which portion of the CA coast is examined and on 
the timescale of analysis. 
The main purpose of this study is to establish a general theory of the mechanism
of interacting Moran effects and to use examples and applications of the theory to provide 
ecological intuition for how the mechanism of interacting Moran effects works. Our goals are
distinct from those of \cite{Kendall2000}, who examined interactions between Moran and dispersal
causes of synchrony.

@@

Our formal theory requires a conceptual understanding of the \emph{spectrum} of an environmental 
or population time series, as well as of the \emph{cospectrum} and \emph{cross spectrum} of two time series,
so we briefly introduce these concepts.
If $w_i(t)$ is a stochastic process or time series measured at location $i$ (e.g., the population density time series
of a species of interest in that location), the spectrum $S_{w_i w_i}(f)$ is a function
of frequency, $f$. For a periodic ocillation, $f$ can be defined as one over 
the timescale, or period, $\sigma$, of the oscillation. 
The spectrum $S_{w_i w_i}(f)$ is larger for frequencies at which $w_i(t)$ oscillates more. So, for example, a 
population $w_i(t)$ that shows strong oscillatory dynamics of $5$-year period will have a large value of 
$S_{w_i w_i}(f)$ for $f=\frac{1}{5} \text{ y}^{-1}$. The spectrum separates the overall variance of a time series into
contributions which occur at different frequencies, in the sense that an integral of $S_{w_i w_i}(f)$ across all 
frequencies equals $\var(w_i)$. In a similar way, the cospectrum of two time series, $w_i(t)$ and $w_j(t)$,
is a function of $f$ that takes large values at frequencies for which oscillations in $w_i(t)$ and 
$w_j(t)$ are both strong and strongly correlated, i.e., they have the same or similar phase. 
Here, $j$ refers to another location where measurements were taken. The cospectrum is 
the real part of the cross spectrum, $S_{w_i w_j}(f)$, which is a complex-valued function of frequency.
The cross-spectrum takes large-magnitude values at frequencies, $f$, for which the oscillatory components
of $w_i(t)$ and $w_j(t)$ are strong and in a consistent phase relationship to each other; 
and the complex phase of the cross spectrum at $f$ then quantifies
that relationship. The notation $S_{ww}(f)$ refers to the \emph{spectral matrix}, 
which has $ij^{\text{th}}$ entry $S_{w_i w_j}(f)$. Spectral methods are standard  
[e.g., \cite{Vasseur2007,Defriez2017}], and many background references are available 
[e.g., \cite{Brillinger2001}].

@@

Our population model is 
\begin{linenomath}
\begin{align}
w_i(t) &= b_1 w_i(t-1)+ \cdots + b_n w_i(t-n) \label{main_model_1} \\ 
&+ p_0^{(1)} \epsilon_i^{(1)}(t)+ \cdots + p_{m_1}^{(1)} \epsilon_i^{(1)}(t-m_1) \label{main_model_2} \\
&+ p_0^{(2)} \epsilon_i^{(2)}(t)+ \cdots + p_{m_2}^{(2)} \epsilon_i^{(2)}(t-m_2) \label{main_model_3} \\
&+ \delta_i(t), \label{main_model_4}
\end{align}
\end{linenomath}
where $w_i(t)$ is an index of the population in location $i=1,\ldots,N$ at time
$t$, and $\epsilon^{(1)}=(\epsilon_1^{(1)},\ldots,\epsilon_N^{(1)})$,
$\epsilon^{(2)}=(\epsilon_1^{(2)},\ldots,\epsilon_N^{(2)})$ and $\delta=(\delta_1,\ldots,\delta_N)$
are environmental processes at the same locations. The processes $\epsilon^{(1)}$ and $\epsilon^{(2)}$ are
taken to be measured, whereas $\delta$ represents the 
aggregate influence of unmeasured processes. We assume that the combined process 
$(\epsilon_1^{(1)},\ldots,\epsilon_N^{(1)},\epsilon_1^{(2)},\ldots,\epsilon_N^{(2)},\delta_1,\ldots,\delta_N)$ is an ergodic second-order stationary stochastic process \citep{Brillinger2001}
with expected values of its components equal to $0$. We make additional mild
regularity assumptions for model stability 
(SI section \ref{SM-sect:model_spec}). This model can be seen as 
a linearization of a very general dynamical model, influenced by ``weak noise'' [see, e.g.,
SI section S1.2 of \cite{Desharnais2018} and SI section S1 of \cite{Walter2017}].
Linearization and ``weak noise'' assumptions have been commonly adopted to make
theoretical progress in ecology, and it has been demonstrated that results based on a
weak noise assumption often hold for noise which is fairly strong \citep{Nisbet1977,Desharnais2018}.
See \cite{Brillinger2001} for background on stochastic processes.

@@

%*DONE relationship to earlier papers comparing synchrony of drivers and pops
It has been a frequent topic of research why populations are often less synchronous, or sychronous over a smaller
spatial extent, than might be expected given the strength and extent of 
synchrony of an environmental driver \citep{Herfindal2022}. Our new mechanism
of interacting Moran effects provides both a new means by which populations may be less synchronous than 
population drivers; and also a new means by which populations can be more 
synchronous than environmental drivers. 
Previously known mechanisms by which populations can be less synchronous than environmental drivers 
include demographic stochasticity and measurement error. 
Antagonistic interactions between Moran drivers may be a common and previously
unrecognized additional mechanism contributing to this discrepancy. On the other hand,
two recent papers descibed ``enhanced Moran effects'' by which specific 
patterns of temporal autocorrelation
in Moran drivers can theoretically cause greater synchrony in populations 
than in drivers \citep{Massie2015,Desharnais2018}.
Synergistic interaction between Moran drivers are another mechanism by which populations can
be more synchronous than expected.

@@

%*Revisit the kendall paper, briefly. Talk about possible three-way interactions.
We have focussed on interactions between two Moran drivers, but synchrony in most systems
may be a phenomenon with multiple (more than two) interacting causes.
\cite{Kendall2000} considered interactions between dispersal and a Moran mechanism of synchrony. Their research
questions were therefore distinct from ours, but combining their viewpoints and ours may lead to 
future work about interactions between dispersal and more than one Moran driver of synchrony. 
Although we considered only two Moran drivers in 
our theory and examples, essentially all population systems are influenced by multiple environmental drivers,
and environmental drivers very commonly are associated with large scale climatic phenomena
such as ENSO, and hence are associated with each other. Thus it may be quite common for synchrony to simultaneously be caused by dispersal and
multiple distinct Moran effects, and these influences may interact in multifarious combinations. 
It may be necessary in future
work to consider interactions between dispersal and two Moran drivers. 
It may be useful to consider cases for which multiple related Moran drivers all interact.
Dispersal can readily be added to our modelling framework:
\cite{Desharnais2018} performed a spectral analysis on a model similar to ours
which included dispersal.
Future work should consider whether and when scenarios of multi-driver interactions between causes of synchrony 
can lead to synchrony patterns which differ fundamentally from what one would expect from one
or two mechanisms. Potential interactions increase as the square of the number of 
drivers, so interactions seem likely to become more important as our viewpoint of synchrony expands.
